---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv") %>%  
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

# Sum coding, see BW:127
contrasts(sim$Sem) <- contr.sum(2)
contrasts(sim$Phon) <- contr.sum(2)

# Define the formula
sim_formula <- bf(Distance ~ Sem * Phon)
                      
# No random effects since there is only one measurement per participant/"item" (?)

# Get priors
get_prior(sim_formula, sim)

my_priors <- c(prior(normal(107, 50), class = "Intercept"), # Got this from 06_homwork
               prior(normal(0, 50), class = "b", coef = "Phon1"), # Where do I get this new name from
               prior(normal(0, 50), class = "b", coef = "Sem1")) # Same

# Am I missing interaction and sigma?
# prior(normal(0, 50), class = "b", coef = "Phon1:Sem1")
# prior(normal(0, 50), class = "sigma")

# Here I suppose that the effect of Sem and Phon are the same, right? Does that make sense?
# I am in general lost here
# How to specify priors, am I missing a plot maybe? What specifically are those numbers based on??
```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

sim_model <- brm(sim_formula, # Just copied this below from 07_hw
                  family = gaussian(),
                  prior = my_priors,
                  data = sim,
                  chains = 4,
                  cores = 4, 
                  iter = 2000,
                  warmup = 1000,
                  thin = 1,
                  seed = 42)
```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
# From 06_hw: Add your priors to the model below. We will run the model and sample only from our priors, i.e. we ignore the data. This is a good sanity check to see if our priors make sense and what they would predict. Plug in the name of your priors from (C) and run the code chunk.

sin_model_prior <- brm(sim_formula,
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            data = sim,
            seed = 42)

sin_model_prior %>%  
  # first extract the samples
  spread_draws(b_Intercept, b_Sem1) |> 
  # Create two columns that represent the two categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_Sem1) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem~., ncol = 1)

sin_model_prior %>%  
  # first extract the samples
  spread_draws(b_Intercept, b_Phon1) |> 
  # Create two columns that represent the two categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_Phon1) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Phon", values_to = "Distance") |> 
  # plot
  ggplot(aes(x = Distance, fill = Phon)) +
    geom_density() +
    facet_wrap(Phon~., ncol = 1)

# Haha, this looks really bad, I look forward to seeing what this is supposed to look like
```

```{r}
#| label: model_fit

# I should probably do this in one but I couldn't figure it out

# From 06_hw: If you are happy with your priors (probably shouldn't be), run the full model and plot the posteriors for semantically different and similar cities just like above. How did the posteriors change after accounting for the data?

sim_model %>% 
  # first extract the samples
  spread_draws(b_Intercept, b_Sem1) %>%  
  # Create two columns that represent the two categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_Sem1) %>% 
  # and then bring into a long format like the data
  select(Different, Similar) %>%
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") %>%  
  # plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem~., ncol = 1)

sim_model %>% 
  # first extract the samples
  spread_draws(b_Intercept, b_Phon1) %>%  
  # Create two columns that represent the two categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_Phon1) %>% 
  # and then bring into a long format like the data
  select(Different, Similar) %>%
  pivot_longer(everything(), names_to = "Phon", values_to = "Distance") %>%  
  # plot
  ggplot(aes(x = Distance, fill = Phon)) +
    geom_density() +
    facet_wrap(Phon~., ncol = 1)

# Interpretation: does this not look okay?
```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

mcmc_plot(sim_model, type = "trace") # looks good, i.e. like caterpillars
mcmc_dens_overlay(sim_model, pars = "b_Sem1") # looks okayyy I guess
# I just guessed what to put in here (b_Sem1)
mcmc_dens_overlay(sim_model, pars = "b_Phon1") # same
sim_model # I think Bulk-ESS is fine (above 400 at least)
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

summary(sim_model)

# Shouldn't I also have some plots
# Not the ones in (c) thought, right?
# I have to stop here for now
```

## (g) Write up your entire analysis:

YOUR WRITE UP HERE

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
