---
title: "session07 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, brms, bayestestR)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

library(tidybayes) # I seemed to need this too
library(see) # This too
```

### Introduction to the problem

We will be working with Quantese again, see `01_exercise.qmd` and the other data resources for a refresher.

```{r}
#| label: load-quantese

quantese <- read_csv("../../data/quantese_naming_sim.csv") 
# add log-transformed reaction time
quantese <- quantese |> 
  mutate(log_rt = log(rt))

# have a look
quantese

```

In an earlier homework, we have assessed the following hypothesis:

*"We hypothesize that log-transformed reaction times are affected by lexical frequency"*

We tried to run the following model that included an appropriate random effect structure but it did not converge with `lme4`:

```{r}
#| label: lmer_quantese

# formulate the corresponding LMER
quant_lmer <- lmer(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID), 
                    data = quantese)
# your receive a warning: "boundary (singular) fit: see help('isSingular')"

```

Now we will re-run this model within the Bayesian framework:

### (A) Specify priors

Check which priors this model needs and specify weakly informative priors.

Remember that weakly informative priors are priors that do not bias the posteriors in any direction, but reasonably constrain the parameter space, i.e. only reasonable values are assumed.

For priors of class `Intercept` and `b`, constrain the parameter space with a wide normal distribution.

```{r}
#| label: get_prior

# Don't I need to first visualize the log_rt distribution, e.g. by making a density plot of it?

quantese %>%
  ggplot(aes(x = log_rt)) +
  geom_density(color = "#D685D5", fill = "#E3BAE2") + # important to have nice colors
  theme_minimal()

# The distributions are somewhat skewed towards the left with a heavy right tale

mdl_formula <- bf(log_rt ~ log_frequency)
get_prior(mdl_formula, quantese)

# Is this wrong? Maybe doesn't make sense if there isn't a perfect normal distribution?

min(quantese$log_rt) # 5.9
max(quantese$log_rt) # 7.8

# Should the distribution then allow all values between 5.9 and 7.8?
# Would we then expect values around 6.85 to be the most common, since it's in the middle of 5.9 and 7.8?
# With values that are lower or higher becoming less common?
# On the density plot, it seems that around 6.35 is the most common
plot_normal(mean = 6.85, sigma = 2) # Should I base this on the density plot or no? It doesnt't do that in 06_homework, I think, so here it's not
# Why does this not work now, it did before??
# The 2 is kind of random

# Specify weakly informative priors

# `Intercept` reflects the average log_rt measure and `log_frequency` reflects the estimated difference between the Intercept and ?? _____ ??

my_priors <- c(prior(normal(6.85, 2), class = 'Intercept'), # 6.85 is in the middle of the min and max log_rt
               # I'm not really sure where the second number is supposed to come from
               prior(normal(0, 2), class = 'b', coef = 'log_frequency')) # 0 because we "pretend to believe" that log_frequency doesn't have an effect at all? 
               # # Same here with the second number - is it always the same as the second one above?

```

We can also now get used to specifying priors for classes `sd` and `cor`. Priors of class `sd` are priors for the variance associated with our random effects. You see here 2 `sd` priors for group `ItemID`. The first one is just a generic prior for all `ItemID` coefficients and the second one with the `coef = Intercept` stands for the random intercept variation between items. Similarly for group `PartID`, we have 3 priors. The first one is again just a generic prior for all `PartID` coefficients. The second one with the `coef = Intercept` stands for the random intercept variation between participants. And the last one with the `coef = log_frequency` stands for the random slope variation between participants for the predictor `log_frequency`. How do we specify these priors. We have usually no idea what to expect from these parameters, so we here also specify very diffuse priors centered on zero. We pick a cauchy distribution (remember? super squashed normal distributions) because they have thick tails. Let's make them super diffuse and pick a `sd = 3` for both items and participants.

```{r}
#| label: add_sd_priors

# add weakly informative priors for sd
my_priors <- c(my_priors, 
               prior(cauchy(0, 3), class = 'sd', group = 'ItemID'),
               prior(cauchy(0, 3), class = 'sd', group = 'PartID'))

```

And finally, if your model as random slopes, they have by default a correlation parameter for intercept and slope. We are not getting into the math here and will follow advice from smart people. We use Lewandowski-Kurowicka-Joe priors (shape = 1): `lkj(1)`. Specifying this prior for random slope models actually helps a lot with the fitting procedure.

```{r}
#| label: add_cor_priors

# add weakly informative priors for cor
my_priors <- c(my_priors, 
               prior(lkj(1), class = 'cor'))

```

```{r}
# EXTRA

# I decided to try this from 06_homework

xmdl_prior <- brm(log_rt ~ log_frequency +
            (1 + log_frequency | PartID) +
            (1 + log_frequency | ItemID),
            family = gaussian(),
            prior = my_priors,
            sample_prior = "only",
            data = quantese,
            seed = 42)

xmdl_prior %>% 
  # first extract the samples
  spread_draws(b_Intercept, b_log_frequency) %>% 
  # Create two columns that represent the two categories
  mutate(log_rt = b_Intercept,
         log_frequency = b_Intercept + b_log_frequency) %>% 
  # and then bring into a long format like the data
  select(log_rt, log_frequency) %>% 
  pivot_longer(everything(), names_to = "log_frequency", values_to = "log_rt") %>% 
  # plot
  ggplot(aes(x = log_rt)) +
  geom_density(color = "#F54927", fill = "#F56F53") +
  theme_minimal()

# I'm not doing facet_wrap since it's not categorical, but is this right? Shouldn't log_frequency come in here somehow?
```

### (B) Run model

Now run the model with 4 chains, 1000 warm-up and 1000 post-warm-up samples per chain. Check that chains are reasonably mixed and that sampling yielded a sufficient effective sample size. If not address the issues appropriately (check slides again).

```{r}
#| label: model_quant

# if the model takes way too long to sample (more than 5-10 mins), use the following subset
# quantese_sub <- quantese |> 
#   filter(PartID %in% c("01","02","03","04",
#                        "05","06","07","08"))

normal_mdl <- brm(log_rt ~ log_frequency +
                  (1 + log_frequency | PartID) +
                  (1 + log_frequency | ItemID),
                  family = gaussian(),
                  prior = my_priors,
                  data = quantese,
                  chains = 4, cores = 4,
                  iter = 2000, warmup = 1000,
                  thin = 1, seed = 42) # I don't know if I need this

normal_mdl

mcmc_plot(normal_mdl, type = "trace") # This looks fine generally, I don't know if no 2-3 in column 1 are a bit sketchy
# Also, what are the reds?
```

### (C) Plot posterior predictions

Find a way to extract posterior samples from the model for the `Intercept` and the `log_frequency` coefficient and plot the predicted posterior mean linear model alongside its 95% Credible Intervals. There are many different ways to get there.

```{r}
#| label: plot_predictions

normal_mdl %>% 
  # first extract the samples
  spread_draws(b_Intercept, b_log_frequency) %>% 
  # Create two columns that represent the two categories
  mutate(log_rt = b_Intercept,
         log_frequency = b_Intercept + b_log_frequency) %>% 
  # and then bring into a long format like the data
  select(log_rt, log_frequency) %>% 
  pivot_longer(everything(), names_to = "log_frequency", values_to = "log_rt") %>% 
  # plot
  ggplot(aes(x = log_rt)) +
  geom_density(color = "#538CF5", fill = "#91B6FA") +
  theme_minimal()
  
# I'm not doing facet_wrap since it's not categorical, but is this right? Shouldn't log_frequency come in here somehow?

# There is some stuff here that might not make sense (we have a continuous variable here, not categorical like in 06_homework) + I have to add 95% Credible Intervals - maybe find it another place? OSF maybe… A bit difficult when the model itself doesn't run correctly

# Based on from 06_homework_solutions:
#xmdl |> 
  # first extract the samples
  #spread_draws(b_Intercept, b_SemSimilar) |> 
  # Create two columns that represent the two categories
  #mutate(Different = b_Intercept,
  #       Similar = b_Intercept + b_SemSimilar) |> 
  # and then bring into a long format like the data
  #select(Different, Similar) |> 
  #pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # plot
  #ggplot(aes(x = Distance, fill = Sem)) +
  #  geom_density() +
  #  facet_wrap(Sem~., ncol = 1)
```

P.S. I didn't have much time to do this and suddenly, I had some issues running code that were completely ok before - I will have to look at this better
